borrow-lambda

### 2025-04-20 paper/paper.tex
last minute fixes

\begin{abstract}
    Borrow checking and ownership is a solution to many of the problems that commonly arise in programming languages with manual memory management like C.
    Like the strategy of garbage collection implemented in several common general purpose programming languages, borrow checking eliminates the
    need for programmers to manually allocate and free the memory their program uses, which is the source of many human errors in programming.
    However, the need for high performance code makes garbage collection a sub-optimal solution, as garbage collection introduces performance
    overheads, non-determinism, and potentially memory fragmentation. Borrow checking and ownership is an alternative that takes a different
    approach - it relies on the strength of the programming language's typesystem. It is a set of rules that allows the compiler to
    automatically determine where to allocate and free memory, both eliminating the need of a garbage collector and the need for a programmer
    to manually allocate and free memory. In this paper, we discuss borrow checking as a solution to common memory errors, and provide
    and implementation on a small extension of the simply-typed lambda calculus.
\end{abstract}

@@

Over time, we have seen that relying on programmers to be skilled enough to avoid
making memory errors in their code is not feasible. When codebases become large scale,
these errors become extremely hard to avoid. Computer viruses and exploits due to
things like buffer overflows affect many people and are still present in a large
amount of software today.

@@

Work has been done to improve manually memory managed languages like C in terms of
safety, by strenghthening the typesystem itself, and attacking the semantics of the
language. Cyclone, a Cornell lab project, developed a dialect of C that works to
ensure memory safety at compile time. Among the features included were statically
checked lifetimes of pointer variables, and restrictions on types of pointers which
work to ensure safety with regards to multi-aliasing \cite{grossman2002region}. Similarly, the
concept of ownership types was conceived as a software design principle, particularly
in research on concurrent programming, where an object governs an area of memory,
and administers any modifications to it \cite{boyapati2002ownership} \cite{gordon2012uniqueness}. In the modern day,
many of these features were materialized in the Rust programming language, a popular
systems-level programming language that is making the case for the assistance of
borrow checking and ownership for high performance code.

@@

        fn borrower<'a>(r: &'a u32) -> &'a u32 {
            let b = 0;
            &b // error: variable 'b' does not live long enough 
        }
    \end{lstlisting}
    \caption{Example of a function with a polymorphic lifetime variable}
    \label{lfpolyex}
\end{figure}
The function \texttt{borrower} requires that the reference passed in, and the
reference returned out, has a lifetime \texttt{'a}. When the function is called,
this variable \texttt{'a} is instantiated to the lifetime of the reference that
is passed in, that being \texttt{\&n}. Thus, the typechecker can determine that
the implementation of the function is invalid, as the reference \texttt{\&b} is
being returned out of the function, but has a lifetime smaller than the lifetime
of \texttt{\&n}, which is what \texttt{'a} was instantiated to.

\section{Implementation}
Our implementation consists of a small extension to the simply-typed lambda calculus,
with constructs that are both necessary and illustrative to borrow checking and ownership,
as well as the typing rules that facilitate borrow checking and ownership. The
implementation is written in OCaml, with each of the constructs encoded as
OCaml types. The main relevant component is the typechecker, which runs in two
phases that will be described in this section. Additionally, we have included a
parser which collects source code into terms represented as nodes in an abstract syntax tree, 
and an evaluator in order to develop a fully functional REPL. The evaluator follows a simple
call by value semantics.

\subsection{Syntax}
Our syntax of terms is as follows.
\begin{center}
    \begin{bnf}
        $t$ : \textsf{Term} ::=
        | $x$ | $\lambda x\ldotp t$ | $t_1\ t_2$ | $\&x$ | $\&\text{mut}\ x$ | $*x$
        | if $t_1$ then $t_2$ else $t_3$
        | let $x = t_1$ in $t_2$
        | $x \coloneq t$
        | $*x \coloneq t$
        | true
        | false
        | unit
        | 0
        | pred $t$
        | succ $t$
        | iszero $t$
        | natvec($t_1,t_2,\dots t_n$)
        | get($t_1,t_2$)
        | getmut($t_1,t_2$)
        | push($t_1,t_2$)
        | pop($t_1$) 
        | $t\ \colon T$
    \end{bnf}
\end{center}
In addition to standard natural numbers and booleans and some simple operations for them, we add
mutable and immutable references, the dereference operator, assignment and
dereference-assignment, as well as a vector of nats for a more complex
(non-copyable) data structure.

@@

@@

    \begin{bnf}
        $\mu$ : \textsf{Mod} ::= mut | shr
    \end{bnf}
\end{center}
Lifetimes are represented by numeric values, corresponding to their scope depth.
We can make this definition, as each binding site introduces a new scope. The
subtyping relation works by then considering a reference with a larger lifetime variable a subtype
of a reference with a smaller lifetime variable. The mod corresponds to the type
of reference (mutable or shared, i.e. immutable). The types of references will be
annotated with both of these constructs, in addition to a type and the variable that they borrow.
Functions can be polymorphic over lifetimes, as denoted by the $\forall \alpha.$ quantifier. This is required for any functions
that have a reference type as a parameter, since the concrete lifetime is not important (or even knowable).

@@

\subsection{Typechecking Pass}
We use bidirectional typechecking, as we felt that this was the simplest and
most natural solution, especially for terms like the if clause whose type depends
on that of the sub terms. As expected, the typechecker checks that terms are
well-typed, and annotates the corresponding abstract syntax tree node with
the determined type of the node, generating a new abstract syntax tree with nodes
annotated with their types.

@@

\begin{figure}[h]
    \begin{lstlisting}[language=caml]
        let a = 0 in        (* a has lifetime '0 *)
        let b = succ 0 in   (* b has lifetime '1 *)
        let c = (\d. d) in  (* c has lifetime '2 *)
        let d = &b in       (* d has type &'1 nat *)
        let e = &mut a in   (* e has type &'0 mut nat *)
        ...
    \end{lstlisting}

@@

\subsection{Borrow Checking Pass}
\begin{figure}

@@

The borrow checking pass is responsible for checking rules related to ownership
and borrowing. Most typing rules of interest are shown in Figure \ref{borrowrules}. Note that
contexts are omitted from a rule if they are not used and don't change in the derivation.
Rule T-Var checks the affine type property - that no value is used more than once.
This is done by maintaining a context of variable usages which we call $\Delta$. Depending on if the type
of the value is copyable, the variable gets added to this set (T-Var). For a copyable variable,
we don't need to track its usage for the purpose of validating move semantics (T-VarCopy).

@@

Additionally, we use the semicolon as a shorthand to denote both a variable's type (on the left) and lifetime (on the right) in
the $\Gamma$ (variables) context together. The rule states that
when a reference is returned out with lifetime $\alpha$, the program typechecks
if the current scope of the closure, which is reflected by the lifetime variable
of bound variable $x$, is smaller than the scope of the reference that is being
returned out from. This means that the reference that is being returned out of
the closure outlives the closure itself, which is what we want. Additionally, when checking
the inner term $t_2$, we add the bound variable $x$ with type $T_1$ and assign it a lifetime index
corresponding to the current scope depth, which is equivalent to the size of the set $\Gamma$.

>>>>

\begin{abstract}
    Borrow checking and ownership is a solution to many of the problems that commonly arise in programming languages with manual memory management like C.
    Like the strategy of garbage collection implemented in several common general purpose programming languages, borrow checking eliminates the
    need for programmers to manually allocate and free the memory their program uses, which is the source of many human errors in programming.
    However, the need for high performance code makes garbage collection a sub-optimal solution, as garbage collection introduces performance
    overhead, non-determinism, and potentially memory fragmentation. Borrow checking and ownership is an alternative that takes a different
    approach - it relies on the strength of the programming language's type system. It is a set of rules that allows the compiler to
    automatically determine where to allocate and free memory, both eliminating the need of a garbage collector and the need for a programmer
    to manually allocate and free memory. In this paper, we discuss borrow checking as a solution to common memory errors, and provide
    an implementation on a small extension of the simply-typed lambda calculus.
\end{abstract}

@@

Over time, we have seen that relying on programmers to be skilled enough to avoid
making memory errors in their code is not feasible. As codebases grow larger and larger,
these errors become extremely hard to avoid. Computer viruses and exploits due to
things like buffer overflows affect many people and are still present in a large
amount of software today.

@@

Work has been done to improve manually memory managed languages like C in terms of
safety, by strengthening the typesystem itself, and attacking the semantics of the
language. Cyclone, a Cornell lab project, developed a dialect of C that works to
ensure memory safety at compile time. Among the features included were statically
checked lifetimes of pointer variables, and restrictions on types of pointers which
work to ensure safety with regards to multi-aliasing \cite{grossman2002region}. Similarly, the
concept of ownership types was conceived as a software design principle, particularly
in research on concurrent programming, where an object governs an area of memory,
and administers any modifications to it \cite{boyapati2002ownership} \cite{gordon2012uniqueness}. In the modern day,
many of these features were materialized in the Rust programming language, a popular
systems-level programming language that is making the case for the assistance of
borrow checking and ownership for high performance code.

@@

        fn borrower<'a>(r: &'a u32) -> &'a u32 {
            let b = 0;
            &b // error: variable 'b' does not live long enough 
        }
    \end{lstlisting}
    \caption{Example of a function with a polymorphic lifetime variable}
    \label{lfpolyex}
\end{figure}
The function \texttt{borrower} requires that the reference passed in, and the
reference returned out, has a lifetime \texttt{'a}. When the function is called,
this variable \texttt{'a} is instantiated to the lifetime of the reference that
is passed in, namely \texttt{\&n}. Thus, the typechecker can determine that
the implementation of the function is invalid, as the reference \texttt{\&b} is
being returned out of the function, but has a lifetime smaller than \texttt{'a}.
This is because any possible instantiation of \texttt{'a} must must be strictly larger than the scope of the function since
it is tied to a parameter.

\section{Implementation}
Our implementation consists of a small extension to the simply-typed lambda calculus,
with constructs that are both necessary and illustrative to borrow checking and ownership. The
implementation is written in OCaml, with each of the constructs encoded as
OCaml types. The main relevant component is the typechecker, which runs in two
phases that will be described in this section. Additionally, we have included a
parser which collects source code into terms represented as nodes in an abstract syntax tree, 
and an evaluator in order to develop a fully functional REPL. The evaluator follows a simple
call by value semantics.

\subsection{Syntax}
In addition to standard natural numbers and booleans and some simple operations for them, we add
mutable and immutable references, the dereference operator, assignment and
dereference-assignment, as well as a vector of nats for a more complex
(non-copyable) data structure.

@@

The syntax of terms is as follows:
\begin{center}
    \begin{bnf}
        $t$ : \textsf{Term} ::=
        | $x$ | $\lambda x\ldotp t$ | $t_1\ t_2$ | $\&x$ | $\&\text{mut}\ x$ | $*x$
        | if $t_1$ then $t_2$ else $t_3$
        | let $x = t_1$ in $t_2$
        | $x \coloneq t$
        | $*x \coloneq t$
        | true
        | false
        | unit
        | 0
        | pred $t$
        | succ $t$
        | iszero $t$
        | natvec($t_1,t_2,\dots t_n$)
        | get($t_1,t_2$)
        | getmut($t_1,t_2$)
        | push($t_1,t_2$)
        | pop($t_1$) 
        | $t\ \colon T$
    \end{bnf}
\end{center}

@@

    \begin{bnf}
        $\mu$ : \textsf{Mod} ::= mut | shr
    \end{bnf}
\end{center}
Lifetimes are represented by numeric values, corresponding to their scope depth.
We can make this definition, as each binding site introduces a new scope. The
subtyping relation works by then considering a reference with a larger lifetime (smaller scope depth) a subtype
of a reference with a smaller lifetime (larger scope depth). The mod corresponds to the type
of reference (mutable or shared, i.e. immutable). The types of references will be
annotated with both of these constructs, in addition to a type and the variable that they borrow.
Functions can be polymorphic over lifetimes, as denoted by the $\forall \alpha.$ quantifier. This is required for any functions
that have a reference type as a parameter, since the concrete lifetime is not important (or even knowable).

@@

\subsection{Typechecking Pass}
We use bidirectional typechecking, as we felt that this was the simplest and
most natural solution, especially for terms like conditionals whose type depends
on that of the sub terms. As expected, the typechecker checks that terms are
well-typed, and annotates the corresponding abstract syntax tree node with
the determined type of the node, generating a new abstract syntax tree with nodes
annotated with their types.

@@

\begin{figure}[h]
    \begin{lstlisting}[language=caml]
        let a = 0 in        (* a has lifetime '0 *)
        let b = succ 0 in   (* b has lifetime '1 *)
        let d = &b in       (* d has type &'1 nat *)
        let e = &mut a in   (* e has type &'0 mut nat *)
        ...
    \end{lstlisting}

@@

\subsection{Borrow Checking Pass}

\begin{figure}
    \begin{center}
        $\Gamma = \cdot \mid \Gamma, x : T ; \alpha$ \quad
        $\Delta = \cdot \mid \Delta, x$ \quad
        $F = \cdot \mid F, x$ \quad
        $B = \cdot \mid B, x_{\mu}$

        \caption{Definitions of typing contexts.}
        \label{contexts}
    \end{center}
\end{figure}

\begin{figure}

@@

The borrow checking pass is responsible for checking rules related to ownership
and borrowing. Most typing rules of interest are shown in Figure \ref{borrowrules}. Note that
contexts are omitted from a rule if they are not used and don't change in the derivation.
Rule T-Var checks the affine type property - that no value is used more than once.
This is done by maintaining a context of variable usages which we call $\Delta$. If the value
is not trivially copyable, the variable gets added to this set (T-Var). For copyable types,
we don't need to track its usage for the purpose of validating move semantics (T-VarCopy).

@@

Additionally, we use the semicolon as a shorthand to denote both a variable's type (on the left) and lifetime (on the right) in
the $\Gamma$ (variables) context together. The rule states that
when a reference is returned out with lifetime $\alpha$, the program typechecks
if the index $\alpha$ is strictly smaller than the current scope. This means that the reference that is being returned out of
the scope outlives the scope itself, which is what we want. Additionally, when checking
the inner term $t_2$, we add the bound variable $x$ with type $T_1$ and assign it a lifetime index
corresponding to the current scope depth, which is equivalent to the size of the set $\Gamma$.
